---
title: "*Smoothing*"
author: "L. Insolia, F. Chiaromonte (special thanks to J. Di Iorio)"
date: "March 24th 2022"
output:
  pdf_document:
    toc: true
  html_document: default
urlcolor: blue
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) # just to clean the pdf files, not recommended in practice!
```


\section{Introduction}

\subsection{Libraries}
We are going to use \textbf{tidyverse} and \textbf{ggplot2}.

```{r}
library(tidyverse)  # for data manipulation and visualization
library(ggplot2)    # for plots
```


\subsection{Data}
We will try to locally regress and smooth the median duration of unemployment (uempmed) based on the \textbf{economics} dataset from \textbf{ggplot2} package. We wil focus on the latest 120 months (10 years from 2005 to 2015)

```{r}
data(economics)
help(economics)
head(economics)
dim(economics)
```

We focus on the latest 120 months.
```{r}

# first note that 
dim(economics)[1] == nrow(economics)

# subset the data
economics <- economics[(nrow(economics)-119):nrow(economics),]
dim(economics)
plot(economics$date, economics$uempmed)
lines(economics$date, economics$uempmed, col='grey60')

```
Transform the dates into indexed from 1 (first measurement in 2005) to 120 (latest measurement in 2015).
```{r}
economics$index <- 1:120
```


\section{LOWESS}
Perform LOWESS in the stats package with the loess command
```{r}
help(loess)
```

Let us focus on the following arguments:
\begin{itemize}
  \item \textbf{formula:} a formula specifying the numeric response and one to four numeric predictors
  \item \textbf{data:} the dataframe
  \item \textbf{span:} the parameter which controls the degree of smoothing
  \item \textbf{degree:} the degree of the polynomials to be used, normally 1 or 2
  \item \textbf{family:} if gaussian fitting is by least-squares, and if symmetric a redescending estimator is used with Tukey's biweight function
\end{itemize}

Let's try different spans and degrees as inputs.
```{r}
loess1_10 <- loess(uempmed ~ index, data = economics, span = 0.1, degree=1) 
loess2_10 <- loess(uempmed ~ index, data = economics, span = 0.1, degree=2) 
plot(economics$uempmed, main="LOESS span=0.1")
lines(predict(loess1_10), col='red')
lines(predict(loess2_10), col='blue')
legend("topright", fill = c("red","blue"), 
       legend = c("degree 1", "degree 2"))

# span=0.25
loess1_25 <- loess(uempmed ~ index, data = economics, span = 0.25, degree=1) 
loess2_25 <- loess(uempmed ~ index, data = economics, span = 0.25, degree=2) 
plot(economics$uempmed, main="LOESS span=0.25")
lines(predict(loess1_25), col='red')
lines(predict(loess2_25), col='blue')
legend("topright", fill = c("red","blue"), 
       legend = c("degree 1", "degree 2"))

# span=0.5
loess1_50 <- loess(uempmed ~ index, data = economics, span = 0.5, degree=1) 
loess2_50 <- loess(uempmed ~ index, data = economics, span = 0.5, degree=2) 
plot(economics$uempmed, main="LOESS span=0.5")
lines(predict(loess1_50), col='red')
lines(predict(loess2_50), col='blue')
legend("topright", fill = c("red","blue"), 
       legend = c("degree 1", "degree 2"))

```

The function loess has the option for fitting the local model through robust $M$-estimator (a generalization of Maximum Likelihood Estimators).
They aresolved through an iterativeley least squares algorithm which trims extreme observations or down-weights the mild ones. 
Specifically, it allows the user to choose the robust Tukey's biweight loss as opposed to the non-robust OLS.

Let's see what the Tukey's biweight loss look alike first, keeping in mind that it is defined as follows:
\begin{itemize} 
  \item Tukey loss: 
    $ \rho(r)= \begin{cases}\frac{c^{2}}{6}\left(1-\left[1-\left(\frac{r}{c}\right)^{2}\right]^{3}\right) & \text { if }|r| \leq c \\ \frac{c^{2}}{6} & \text { otherwise }\end{cases} $ 
  \item Tukey loss derivative: 
    $ \psi(r) = \rho^{\prime}(r)= \begin{cases}r\left[1-\left(\frac{r}{c}\right)^{2}\right]^{2} & \text { if }|r| \leq c \\ 0 & \text { otherwise }\end{cases} $
  \item Tukey loss weight function: 
    $ w_{\psi}(r) =  \frac{\psi(r)}{r} = \begin{cases} \left[1-\left(\frac{r}{c}\right)^{2}\right]^{2} & \text { if }|r| \leq c \\ 0 & \text { otherwise }\end{cases} $
\end{itemize}
```{r, fig.width = 12, fig.height = 6}

library(ggpubr) # to create subpanels within a figure

tukey_loss <- function(r, c) {
  ifelse(abs(r) <= c,
         c^2 / 6 * (1 - (1 - (r / c)^2)^3),
         c^2 / 6)
}
tukey_loss_derivative <- function(r, c) {
  ifelse(abs(r) <= c,
         r * (1 - (r / c)^2)^2,
         0)
}


r <- seq(-6, 6, length.out = 301)
c <- 1:3

# plot of tukey loss
library(ggplot2)
theme_set(theme_bw())
loss_df <- data.frame(
  r = rep(r, times = length(c)),
  loss = unlist(lapply(c, function(x) tukey_loss(r, x))),
  c = rep(c, each = length(r))
)
p1 <- ggplot(loss_df, aes(x = r, y = loss, col = factor(c))) +
  geom_line(size=1) +
  labs(title = "Plot of Tukey loss", y = "Tukey loss",
       col = "c") +
  theme(legend.position = "bottom")

# plot of tukey loss derivative
loss_deriv_df <- data.frame(
  r = rep(r, times = length(c)),
  loss_deriv = unlist(lapply(c, function(x) tukey_loss_derivative(r, x))),
  c = rep(c, each = length(r))
)
p2 <- ggplot(loss_deriv_df, aes(x = r, y = loss_deriv, col = factor(c))) +
  geom_line(size=1) +
  labs(title = "Plot of derivative of Tukey loss", y = "Derivative of Tukey loss",
       col = "c") +
  theme(legend.position = "bottom")

# plot of tukey loss weight function
p3 <- ggplot(loss_deriv_df, aes(x = r, y = loss_deriv/r, col = factor(c))) +
  geom_line(size=1) +
  labs(title = "Plot of weight function of Tukey loss", y = "Weight function of Tukey loss",
       col = "c") +
  theme(legend.position = "bottom")

figure <- ggarrange(p1, p2, p3,
                    labels = c("A", "B", "C"),
                    ncol = 3, nrow = 1)
figure


# compare it with OLS!
p1ols <- ggplot(loss_df, aes(x = r, y = r^2/2)) +
  geom_line(size=1) +
  labs(title = "Plot OLS loss", y = "OLS loss",
       col = "c")

p2ols <- ggplot(loss_df, aes(x = r, y = r)) +
  geom_line(size=1) +
  labs(title = "Plot OLS loss derivative", y = "OLS loss derivative",
       col = "c")

p3ols <- ggplot(loss_df, aes(x = r, y = r/r)) +
  geom_line(size=1) +
  labs(title = "Plot OLS weight function", y = "OLS weight function",
       col = "c")

figure <- ggarrange(p1ols, p2ols,p3ols,
                    labels = c("A", "B", "C"),
                    ncol = 3, nrow = 1)
figure

```



To use this option in loess, we use the argument family="symmetric".
```{r}
# span=0.25
loess1_25sim <- loess(uempmed ~ index, data = economics, span = 0.25, 
                      degree=1, family="symmetric")

# compare it with the default OLS
plot(economics$uempmed, main="LOESS span=0.25")
lines(predict(loess1_25), col='red')
lines(predict(loess1_25sim), col='green',)
legend("topright", fill = c("red","green"), 
       legend = c("OLS degree 1", "Tukey degree 1"))

```

Note that ggplot uses loess in its geom_smooth function
```{r}
economics %>% ggplot(aes(index,uempmed)) + geom_point() + geom_smooth()
```

We should be careful for the default parameters used in geom_smooth, but we can change them as follows:
```{r}
economics %>% ggplot(aes(index,uempmed)) + geom_point() + 
  geom_smooth(method="loess", span=0.15, methods.args = list(degree=1))
```

\section{Bin Smoothing and Kernel Smoothing}

The general idea of smoothing is to group data points into strata in which the associated trend changes "slowly". 
For this reason we can assume the trend to be constant within a small window. 
In our case we can assume that the unemployment remains approximately the same within a given 3-months' time window.

The assumption implies that the average of the values in the window (in this case: 3 months) provides a good estimate. 
By computing this mean for every point (moving the window), we obtain a smooth estimate of the underlying curve.

The command that we are going to use is \textbf{ksmooth}.
```{r}
help(ksmooth)
```

If the mean is computed giving equal weights to the points belonging to each window, we talk about "box" kernel.
The result is a list with the original \textbf{x} and the new smoothed values \textbf{y}.
```{r}
window <- 3
box_smooth <- ksmooth(economics$index, economics$uempmed, kernel='box', bandwidth = window)
box_smooth
```

Let's plot our result using ggplot (unlike base R plots).
```{r}
economics %>% mutate(smooth = box_smooth$y) %>% ggplot(aes(index, uempmed)) + 
  geom_point(size=3, alpha=0.5, color='grey') + geom_line(aes(index, smooth), color='red')
```

The result from the bin box smoother is quite wiggly. 

The reasons for this can be the bandwidth (too small) or the uniform weights. 
We can change units' weights by giving larger weights to the ones in the "central" portion of the window, thus the points at the edges will receive very little weights.

Here we will use a weighted average, where the weights are provided by a normal density.
```{r}
norm_smooth <- ksmooth(economics$index, economics$uempmed, kernel='normal', bandwidth = window)

economics %>% mutate(smooth = norm_smooth$y) %>% ggplot(aes(index, uempmed)) + 
  geom_point(size=3, alpha=0.5, color='grey') + geom_line(aes(index, smooth), color='red')
```

It is still wiggly! We need to change the bandwidth.
```{r}
window <- 6 #6 month's time
norm_smooth <- ksmooth(economics$index, economics$uempmed, kernel='normal', 
                       bandwidth = window)

economics %>% mutate(smooth = norm_smooth$y) %>% 
  ggplot(aes(index, uempmed)) + 
  geom_point(size=3, alpha=0.5, color='grey') + 
  geom_line(aes(index, smooth), color='red')

```

\section{General Idea on Kernel Density Estimator}

Let's simulate a new dataset, containing gender (as M/F) and weight of 400 subjects living on an undefined region of the universe.
```{r}
set.seed(1234)
df <- data.frame(
  sex=factor(rep(c("F", "M"), each=200)),
  weight=round(c(rnorm(200, mean=55, sd=5),
                 rnorm(200, mean=65, sd=5)))
)
head(df)
tail(df)
```

Let's produce a histogram and its density through ggplot.
```{r}
ggplot(df, aes(x=weight)) + 
  geom_histogram(aes(y=..density..), colour="black", fill="grey")+
  geom_density(alpha=.2, fill="#FF6666") 
```

We can adjust the default density through the \textbf{adjust} argument (default is 1).

What is the adjust input used for?
```{r}
ggplot(df, aes(x=weight)) + 
  geom_histogram(aes(y=..density..), colour="black", fill="grey")+
  geom_density(alpha=.2, fill="#FF6666", adjust=1/2) 

ggplot(df, aes(x=weight)) + 
  geom_histogram(aes(y=..density..), colour="black", fill="grey")+
  geom_density(alpha=.2, fill="#FF6666", adjust=2) 
```