---
title: "Supervised Dimensional Reduction"
author: "J. Di Iorio, F. Chiaromonte"
date: "3/29/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\section{Libraries}
We are going to use some new and old libraries.

```{r}
library(tidyverse) # for data manipulation and visualization
library(plotly) # for plots in 3D
library(ggplot2) # for plots in 2D
library(mvtnorm) #to generate multivariate normal distribution
library(MASS)
library(dr) # for SIR
library(factoextra)
```

\section{Data}
Let’s first define a function to generate a Gaussian. This function takes four arguments:
\begin{itemize}
\item n: number of observations;
\item center: the mean of the Gaussian
\item sigma: the covariance matrix
\item label: the cluster label
\end{itemize}
```{r}
generateGaussianData <- function(n, center, sigma, label) {
  data = rmvnorm(n, center, sigma)
  data = data.frame(data)
  names(data) = c("x", "y", "z")
  data = data %>% mutate(class=factor(label))
  data
}
```

Now let's simulate a dataset. We plot the data
```{r}
# cluster 1
n = 200
center = c(2, 8, 6)
sigma = matrix(c(1,0.88,0.88,0.88, 1,0.88,0.88,0.88, 1), nrow = 3, byrow=T)
group1 = generateGaussianData(n, center, sigma, 1)
  
# cluster 2
n = 200
center = c(4, 8, 6)
sigma = matrix(c(1,0.88,0.88,0.88, 1,0.88,0.88,0.88, 1), nrow = 3, byrow=T)
group2 = generateGaussianData(n, center, sigma, 2)

# cluster 3
n = 200
center = c(6, 8, 6)
sigma = matrix(c(1,0.88,0.88,0.88, 1,0.88,0.88,0.88, 1), nrow = 3, byrow=T)
group3 = generateGaussianData(n, center, sigma, 3)
  
# all data
df = bind_rows(group1, group2, group3)
```

Now let's plot our simulated data.
```{r}
library(plotly)
fig <- plot_ly(df, x = ~x, y = ~y, z = ~z, color = ~class, colors = c('#b3e378', '#81e5f0', '#ed5391'))
fig <- fig %>% add_markers()
fig <- fig %>% layout(scene = list(xaxis = list(title = 'x'),
                                   yaxis = list(title = 'y'),
                                   zaxis = list(title = 'z')))

fig
```

\section{LDA vs PCA}
\subsection{LDA}
Let's perform LDA:
```{r}
lda.df <- lda(factor(class) ~ x + y + z, data = df)
lda.df
```

Let us plot the projections on LD1 and LD2
```{r}
# prefiction on df to get projections
predmodel.lda = predict(lda.df, data=df)

# projections with original class 
newdata <- data.frame(type = df$class, lda = predmodel.lda$x)
library(ggplot2)
ggplot(newdata) + geom_point(aes(lda.LD1, lda.LD2, colour = type), size = 2.5)

```

\subsection{PCA}
Now let us perform PCA.
```{r}
pc <- prcomp(df[,c(1,2,3)])
get_eig(pc)

```
This is the corresponding biplot: just considering the first principal component it is impossibile to notice differences within the three groups (all the groups are overlapped).
```{r}
fviz_pca_biplot(pc, col.var= "#2E9FDF", col.ind= df$class, label="var")
ggplot(newdata) + geom_point(aes(lda.LD1, lda.LD2, colour = type), size = 2.5)
```

\section{SIR}

Now we use the SIR (Sliced Inversion Regression) in the dr package
```{r}
dr_res <- dr(class ~ x + y + z, data = df, method='sir')
plot(dr_res, col=df$class)
```

```{r}
names(dr_res)
```



